AHABus \& FCORE:\\Modular Bus for High-Altitude Balloon Missions

### 2017-04-10 chapters/03-methodology.tex
Add bus frequency

The following chapter describes the design processes used to develop and
integrate those components into a functioning prototype of the Abertay
High-Altitude Bus (\textit{AHABus}), and the methods used to test the viability
and performance of the design.

@@

\section{Development Approach}

@@

\subsubsection{Initial Testing}

@@

@@

@@

>>>>

The following chapter describes the design processes used to develop and
integrate those components into a functioning prototype of the
\acrfull{ahabus}, and the methods used to test the viability and performance of
the design.

@@

The proposed solution was designed to run on low-cost embedded platforms, and
the prototype was built around the Espressif ESP8266 micro-controller
(\acrlong{obc}). The radio link used a Radiometrix NTX2B transmitter
(\cite{radiometrix2012}), and an unnamed Software-Defined Radio receiver and
antenna.

The software packages used during the project were:

\item[ESP OpenRTOS SDK] A FreeRTOS-based real-time kernel for the ESP8266.
\item[Clang/LLVM] C compiler for macOS (version 802.0.38).
\item[GQRX] macOS Software-Defined Radio application.
\item[dl-fldigi] digital radio decoding software.
\item[Textmate] Code editor.
\item[C (C99 standard)] Language used for all parts of \acrshort{ahabus}.
\item[Python 2.7] Language used for log file analysis

@@

The decoding functions of the library were designed with threaded applications
in mind: the decoding function is given a callback called whenever a packet
is decoded, and never returns.

Decoding is done in a sequential way. Once the start of a frame has been
detected, the next 256 bytes are collected. Those 256 bytes are then checked
at increasing levels of details. Once a frame has been received, the first
action that the decoder runs is \acrlong{fec}. If all errors can be corrected,
the packet header is then extracted from the frame, as well as the beginning of
the packet's data. Subsequent frames are decoded the same way, except all of
their content is added to the packet's data, until the enough bytes have been
decoded to account for the packet's indicated length.

The user-provided callback function is invoked as soon as a full packet is
decoded. It is given a flag indicating whether any erroneous frames were
detected, so that potentially corrupted packets can be marked.

@@

The radio subsystem was designed around the ESP8266's \texttt{UART1}. While it
provides a hardware transmission buffer, tis size is only 127 bytes. Given the
radio channel's low bandwidth (\SI{200}{bps}), this was considered insufficient.
To allow whole packets to be sent to the radio without forcing the calling code
to wait until the buffer is empty, the hardware buffer was supplemented with
a much larger buffer in software. Whenever data is sent to the radio system,
it is placed in this buffer. An Interrupt Service Routine so that whenever the
hardware buffer is empty, 127 bytes are copied from the software buffer to the
hardware one.

To avoid overhead, the radio subsystem was not given any threading protections.
Instead, calling code was designed so that no two places can be interacting with
the radio subsystem at the system: any function that creates packets and places
them in the radio queue is marked as critical, preventing FreeRTOS to switch
contexts.

@@

Because \acrshort{fcore} was designed to run on embedded platforms with 
potentially no support for configuration files, the settings for a given mission
(payload addresses, priority levels...) had to be written as part of the
firmware source code. To avoid requiring future users having to understand and
modify the source code themselves, a code generation approach was taken: a
tool parses a configuration file and generates C source files containing the
mission configuration data, after which the firmware can be compiled and flashed
without manual editing of its source. An example configuration file is shown
in listing \ref{lst:conf-file}

\begin{lstlisting}[
    caption=FCORE - Example Configuration File,
    label=lst:conf-file
]
{
    "name": "FCORE Test 1",
    "payloads": [
        {
            "address":  10,
            "name":     "zero-mass spectrometer",
            "priority": 2
        },
        {
            "address":  20,
            "name":     "thermometer",
            "priority": 1
        }
    ]
}
\end{lstlisting}

This tool was also used to optimise the polling frequency for each payload.
A payload's priority level defines how often it is polled: level two payloads
are polled half as often as level one, and level three a third as often. Since
the number of payloads, their level and the available bandwidth are known when
the file is parsed, a formula could be obtained to calculate the optimal
polling interval, provided the maximum size of a packet was fixed (during the
tests, this value was fixed to 420 bytes, which fit in two 256-byte frames,
resulting in at most 512 bytes generated each time a payload is polled).

If $B$ is the total available bandwidth, $n_1, n_2, n_3$ the number of payloads
at each level, $F_P$ the average byte rate of a level 1 payload and $F_S$ the
average byte rate of the system health packets:

\[B = F_S + n_{1}F_{P} + \frac{n_{2}F_{P}}{2} + \frac{n_{3}F_{P}}{3} \]

\[B = F_S + F_{P}(n_{1} + \frac{n_2}{2} + \frac{n_3}{3}) \]

\[F_P = \frac{B-F_S}{n_{1} + \frac{n_2}{2} + \frac{n_3}{3}} \]

Once the average byte rate of a payload, its priority level $P$ and how many
frames are required to encapsulate the data it can return, the interval can be
obtained:

\[ \Delta_T = P \times \frac{256 \times N_{Frames}}{F_P} \]

The configuration tool was written in ruby because it allows easy parsing of
JSON files. An example output source file is shown in figure \ref{fig:conf-out},
generated from the example configuration file discussed previously.

\begin{lstlisting}[
    caption=FCORE - Example Configuration Output,
    label=lst:conf-out
]
// FCORE Configuration File - FCORE Test 1
// Generated on 2017-03-29 by fcore_config.rb

#define FCORE_MISSION_NAME    "FCORE Test 1"
#define FCORE_CONFIG_GENDATE  "2017-03-29"
#define FCORE_PAYLOAD_COUNT   (2)
#define FCORE_BUS_MAXDATA     (420)
#define FCORE_BUS_FREQ        (10.164040404040405)
#define FCORE_BUS_INTERVAL    (51)

const FCPayload fcore_payloads[FCORE_PAYLOAD_COUNT] = {
    {10, "zero-mass spectrometer", 2, STATUS_BOOT},
    {20, "thermometer", 1, STATUS_BOOT},
};
\end{lstlisting}

@@

Because a large part of the project was focused on developing the protocols and
software that support the \acrshort{ahabus} platform, most of the testing was
done on development boards and breadboards. Because the NTX2 radio transmitter
requires voltage differentials to control the frequency, and it was controlled
using a \acrshort{uart} signal, a voltage divider bridge was added between the
digital I/O pin and its frequency control pin.

\begin{figure}[H]
\includegraphics[width=0.4\textwidth]{ahabus-schema.png}
\centering
\caption{AHABus - Electronic Circuit}
\label{fig:ahabus-schema}
\end{figure}

Some preliminary printed circuit board layouts were designed for a next round
of testing. The schematics and the PCB layout are shown in figures
\ref{fig:ahabus-schema} and \ref{fig:ahabus-pcb}.

\begin{figure}[H]
\includegraphics[width=0.4\textwidth]{ahabus-pcb.png}
\centering
\caption{AHABus - Printed Circuit Board Layout}
\label{fig:ahabus-pcb}
\end{figure}
